---
title: "Mortality in South Africa"
runtime: shiny
output: 
  flexdashboard::flex_dashboard:
    theme: lumen
    orientation: columns
    vertical_layout: fill
---

```{r setup, include=FALSE}
library(flexdashboard)
library(southafricastats)
library(dplyr)
library(ggplot2)
library(tidyr)
library(leaflet)
library(DT)


totals <- population_zaf %>%
          filter(year == 2013) %>%
          select(province, total)

#mutate makes a new column

compare_provinces <- mortality_zaf %>%
    left_join(totals) %>%
    mutate(rate = deaths / total * 1e3) %>%
    group_by(province, indicator) %>%
    summarize(rate = mean(rate)) %>%
    ungroup %>%
    left_join(totals) %>%
    spread(indicator,rate)

mortality_zaf <- mortality_zaf %>%
      filter(indicator != "All causes")




```

Column {.sidebar}
====================================================

This app explores the mortality data from the 
[South Africa Data Portal](http://southafrica.opendataforafrica.org/)

Use input below to select a cause of death to explore


```{r}
selectInput(inputId = "indicator",
            label = "Choose a cause of death:",
            choices = unique(mortality_zaf$indicator),
            selected = "Diabetes mellitus (E10-E14)")


```




Map {data-icon="fa-map-marker"}
====================================================

### Where is the mortality rate highest?

```{r}
selected_df <- reactive({ 
        subset_df <- compare_provinces[,c(1,which(colnames(compare_provinces) == input$indicator))]
        colnames(subset_df) <- c("province","indicator")
        population_zaf %>%
              filter(year == 2013) %>%
              left_join(subset_df, by="province") %>%
              mutate(indicator = indicator / sum(indicator), na.rm=TRUE)
  })

renderLeaflet({
    leaflet(selected_df()) %>%
              addProviderTiles("CartoDB.Positron") %>%
              addCircles(lng = ~longitude,
                         lat = ~latitude,
                         radius = ~sqrt(indicator) * 3e5,
                         weight = 2.5,
                         color = "magenta",
                         popup = ~province)
  
})

```

Comparing provinces {data-icon="fa-list"}
=====================================================

### How does the selected mortality rate compare to the overall mortality rate?


```{r}

scatterplot_df <- reactive({
        subset_df <- compare_provinces[,c(1,2,
                                          which(colnames(compare_provinces) == "All causes"),
                                          which(colnames(compare_provinces) == input$indicator)
                                          )]
        colnames(subset_df) <- c("province","total","x","y")
        subset_df
})

renderPlot({
        ggplot(scatterplot_df(),
               aes(x, y, 
                   label=province)) +
        geom_point(aes(size=total),
                   alpha = 0.8, color = "midnightblue") + 
        geom_text(vjust = 2.5) +
        theme_minimal() + 
          labs(x = "All causes",
               y = input$indicator)
})

```


Changes in time{data-icon="fa-area-chart"}
=====================================================

### How have the number of deaths changed over time?

```{r}

renderPlot({
  mortality_zaf %>%
    filter(indicator == input$indicator) %>%
    ggplot(aes(year, deaths, color = province)) + 
    geom_line(alpha = 0.8, size= 1.5) +
    theme_minimal(base_size = 18) + 
    labs(x = NULL, y="Number of deaths")
  
  
})


```


Table {data-icon="fa-table"}
====================================================

Check out the data here

```{r}
renderDataTable({
  datatable(mortality_zaf %>%
    filter(indicator == input$indicator)
    #,options = list()
    )
  
})

```



